---
title: "506 Group Project: Bootstrapping Regression"
subtitle: '*Analysis of Factors Affecting the Drinking Alcohol Habits in a National Sample of US Adults*'
author: " Group 9: Daxuan Deng, Ming-Chen Lu, Ningyuan Wang"
date: "December 5, 2019"
output: 
  html_document:
  toc: true
  number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction (Ningyuan)



```{r cars}

```

## Data (Ningyuan)



```{r pressure, echo=FALSE}

```



## Method (Daxuan)

Empirical bootstrap is a resampling method for assessing uncertainty of estimator. It is typically used when the exact or asymptotic analytic solutions are unavailable or unsatisfactory. 

Generally, we draw samples with replacement from data set, and calculate the statistic we are interested in. Then, we repeat it $B$ times to generate the empirical distribution of estimator. With this distribution we could derive the variance of estimator.

For example, suppose the regression model is:
$$y \sim X\beta$$
where $y\in R^n$ is the response, $X$ is the $n \times (p+1)$ design matrix, and $\beta \in R^{p+1}$ is the coefficient. In normal setting, we assume that error are Gaussian. Under this assumption, we calculate sample variance of coefficient estimator $Var(\hat{\beta})$ using residual sum of squares (RSS):
$$\widehat{Var} (\hat{\beta}) = {\hat{\sigma}}^2(X'X)^{-1}$$
$${\hat{\sigma}}^2 = \frac{RSS}{n-p-1}$$
However, this could lead a bad result if data is skewed, because the outliers will force the ${\hat{\sigma}}$ become much bigger. To address this problem, we may use the residual bootstrap.

After running regression, we obtain coefficient estimate $\hat{\beta}$ and residuals:
$$r = y  - \hat{y} = y - X\hat{\beta}$$
which is also a vector in $R^n$. Then we sample $r_i^*$ from $\{r_1, r_2,...r_n\}$ with replacement for $i \in \{1,2,...,n\}$, and define
$$r^* = (r_1^*, r_2^*,...,r_n^*)'$$
$$y^* = X\hat{\beta} + r^*$$
In other word, we fix design matrix $X$, but generate a new response vector $y^*$ using the fitted value and the 'error' from resampling the residuals. Now we fit a new model:
$$y^* \sim X\beta$$
and we get a new coefficient estimate $\beta^*$. Repeating this process for $B$ times, we get $B$ bootstrap samples of $\hat{\beta}$:
$$\beta^{*(1)},\beta^{*(1)},...\beta^{*(B)}$$
Finally, we could calculate the sample variance of $\hat{\beta}$:
$$\bar{\beta}^* = \frac{1}{B}\sum_{i=1}^B \beta^{*(i)}$$
$$\widehat{Var}_B(\hat{\beta}) = \frac{1}{B}\sum_{i=1}^B (\beta^{*(i)}-\bar{\beta}^*)^2$$
Basically, that's the core analysis method we use in this project. To do parallel computation, we use different softwares: MATLAB 2018b, Python 3.6.9 and R version 3.6.1 to calculate respectively. Besides, in all the analysis, a two-tailed $P$ value less than 0.05 $(P < 0.05)$ was considered statistically significant.

## Core Analysis (Please follow the parallel orders) {.tabset .tabset_fade}

### R

```{r, eval=FALSE}
#library(reticulate)
#use_virtualenv(virtualenv = "r-reticulate")
1+2
```

### Python

```{python, eval = FALSE, python.reticulate = FALSE}

#import pandas as pd
#import numpy as np
```

### Matlab

```{matlab, eval = FALSE, include = TRUE}
% Title:  Matlab script for group project, STATS 506
% Author: Daxuan Deng
% Funtion: explore the relationship between health condition and drinking
%          alcohol, using data NHANES 2005-2006 data.
% Date: 11/30/2019

% load data
alq = xptread('ALQ_D.XPT');
demo = xptread('DEMO_D.XPT');
hsq = xptread('HSQ_D.XPT');

% select variables
alq = alq(:, {'SEQN', 'ALQ120Q'});
demo = demo(:, {'SEQN', 'RIAGENDR', 'RIDAGEYR', 'DMDEDUC2', 'INDFMPIR'});
hsq = hsq(:, {'SEQN', 'HSD010'});

% merge data
dt = join(alq, demo, 'Keys', 'SEQN');
dt = join(dt, hsq, 'Keys', 'SEQN');

% rename columns
dt.Properties.VariableNames = ...
["id", "alcohol", "sex", "yr", "edu", "pir", "health"];

% drop invalid values
dt = rmmissing(dt);
dt(dt.alcohol > 365, :) = [];
dt(dt.yr < 21, :) = [];
dt(dt.edu > 5, :) = [];
dt(dt.health > 3, :) = [];

% centralize and factorize
dt.alcohol = (dt.alcohol - mean(dt.alcohol)) ./ std(dt.alcohol);
dt.sex = categorical(dt.sex);
dt.yr = (dt.yr - mean(dt.yr)) ./ std(dt.yr);
dt.edu = categorical(dt.edu);
dt.pir = (dt.pir - mean(dt.pir)) ./ std(dt.pir);
dt.health = categorical(dt.health);
% set 'Good' as base level
dt.health = reordercats(dt.health, {'3','1','2'});

% run OLS
md = fitlm(dt, 'alcohol ~ sex + yr + edu + pir + health');

% summary
md.Coefficients

% extract fitted values and residuals 
fit = predict(md, dt(:, 3:7));
res = md.Residuals.Raw;
coef = md.Coefficients(:,1);

% plot residuals
plot(1:height(dt), res, 'x'), title('residuals of OLS')

% bootstrap
rng(506);
nboot = 1000;

% resample residuals
func = @(x)x;
res_boot = bootstrp(nboot, func, res);

dt_boot = dt(:, 3:7);
beta_boot = zeros(nboot, 10);

for i=1:nboot
    % generate new response
    dt_boot.alcohol = fit + res_boot(i,:)';
    
    % fit new model
    md_boot = fitlm(dt_boot, 'alcohol ~ sex + yr + edu + pir + health');
    
    % extract new estimate
    beta_boot(i,:) = table2array(md_boot.Coefficients(:,1))';
end

% hist health_1 and health_2
hist(beta_boot(:,9)),title('coefficient of health 1')
hist(beta_boot(:,10)),title('coefficient of health 2')

% calculate std err
se = std(beta_boot);

% summary
result = coef;
result.se = se';
result.t = result.Estimate ./ result.se;
result.pvalue = 1-tcdf(abs(result.t),1);

result
```


## Additional Analysis (Ming-Chen)


## Results (Brief by Ningyuan )

## Discussion (??? B-Spline)


## References （Ningyuan)
http://faculty.washington.edu/yenchic/17Sp_403/Lec6-bootstrap_reg.pdf
http://users.stat.umn.edu/~helwig/notes/boot-Notes.pdf


